---
title: "Constructing the MPI"
author: "Nate Kratzer"
date: "May 13, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##The Overall Structure

A multidimensional poverty index attempts to capture a more accurate measurement of how poverty is actually experienced. Traditionally, poverty has been measured solely by income. While this may be a decent proxy in some cases, it does mean giving zero weight to other important parts of life. Multidimeensional poverty indexes have become widely used in international development. The Oxford Poverty and Human Development Initiative sums up the reasons for switching from income-based to multidimensional measures of poverty: "empirically, income-poor households are (surprisingly) not well-matched to household carrying other basic deprivations like malnutrition; also the trends of income and non-income deprivations are not matched, and nor does growth ensure the reduction of social deprivations. And, a dashboard overlooks the interconnection between deprivations, which people experience and policies seek to address." (Alkire et al 2015)


This document shows how to construct a multidimensional poverty index for the United States using Census data from the American Community Survey and code in the R programming language. 

The index uses nine indicators across three dimensions

1. Economic
    + Income Poverty
    + Employment
    + Housing Costs
2. Educational
    + Educational Attainment
    + Internet Access
    + Linguistic Isolation
3. Health
    + Insurance
    + Disability
    + Overcrowding

##The initial data

Data comes from IPUMS USA and covers 2013, 2014, and 2015. I have commented out the lines of code used to read in the data from the original downloaded csv. After loading, I suggest saving in R's feather format to save time. 

```{r data, message = FALSE}
library(survey)
library(stargazer)
library(tidyverse)
library(feather)

#data = read_csv("C:/Users/natek/Documents/usa_00007.csv")
#write_feather(data, "C:/Users/natek/Documents/MPI feather.feather")

data = read_feather("C:/Users/natek/Documents/MPI feather.feather")

str(data)

```

## Economic

#### Income Poverty

The income poverty indicator is straightforward and is the same for all members of the household. The ACS data already contains a poverty indicator showing percent of poverty line for all households. I recode this as a binary variable such that households are poor (1) or not poor (0)

```{r income}
data$income_poverty = if_else(data$POVERTY < 100, 1, 0)
```

#### Family Employment

Family employment is also assigned at the household level. If no one in the family is employed, then the family is initially considered employment deprived. However, if anyone in the family has retirement income, or investment income in excess of $25,000, or if the family lives on a working farm, then the family is not considered employment deprived. 

```{r employment, message = FALSE}
data$INCINVST[data$INCINVST == 999999] <- 0
data$INCRETIR[data$INCRETIR == 999999] <- 0

data = data %>%
  mutate(emp_num = if_else(EMPSTAT == 1, 1, 0)) %>%
  group_by(SERIAL, YEAR) %>%
  mutate(retired = if_else(any(INCRETIR > 1), 1, 0),
         investment = if_else(any(INCINVST > 25000), 1, 0),
         fam_emp = if_else(any(emp_num ==1)|any(retired ==1)|any(investment == 1)|FARM == 2, 0, 1))

```

## Health

#### Health Insurance

Health insurance is assigned at the individual level. Some members of the household may have health insurance, while others do not. 
```{r health, message = FALSE}
data$health = if_else(data$HCOVANY == 2, 0, 1)

```

#### Disability

Disability, for this index, is defined as those who said either they had difficulty caring for themselves or independent living difficulty. 
```{r disability, message = FALSE}
data$disability <- if_else(data$DIFFMOB == 2 |data$DIFFCARE == 2, 1, 0)
```


## Education

Education deprivation is at the individual level. It is calculated differently for those under the age of 19. For those 19 and older, a lack of high school degree is considered educationally deprived. For those under 18, being in a grade lower than one's age warrants is considered to be educationally deprived. For example, a 7-year-old who is attending kindergarten or an 18-year-old who is attending 11th grade. 

```{r education, message = FALSE}
data$education = if_else(data$EDUCD < 62 & data$AGE > 18 & data$EDUCD !=1, 1, 0) #less than hs, over 18, not NA

# data$grade_recode = NA ##where GRADEATTD == 0, it stays NA, everywhere else it changes. 0 is NA in GRADEATTD codebook
# data$grade_recode[data$GRADEATTD < 30 & data$GRADEATTD != 0] = 0
# data$grade_recode[data$GRADEATTD == 30] = 4
# data$grade_recode[data$GRADEATTD == 31] = 1
# data$grade_recode[data$GRADEATTD == 32] = 2
# data$grade_recode[data$GRADEATTD == 33] = 3
# data$grade_recode[data$GRADEATTD == 34] = 4
# data$grade_recode[data$GRADEATTD == 40] = 8
# data$grade_recode[data$GRADEATTD == 41] = 5
# data$grade_recode[data$GRADEATTD == 42] = 6
# data$grade_recode[data$GRADEATTD == 43] = 7
# data$grade_recode[data$GRADEATTD == 44] = 8
# data$grade_recode[data$GRADEATTD == 50] = 12
# data$grade_recode[data$GRADEATTD == 51] = 9
# data$grade_recode[data$GRADEATTD == 52] = 10
# data$grade_recode[data$GRADEATTD == 53] = 11
# data$grade_recode[data$GRADEATTD == 54] = 12
# data$grade_recode[data$GRADEATTD > 59] = 120
# 
# 
# data$education[data$AGE < 19 & ((data$GRADEATTD+6) < data$AGE)] = 1 ##If age is more than 6 years above grade level

data = data %>%
  group_by(SERIAL, YEAR) %>%
  mutate(fam_edu = if_else(any(EDUCD > 80), 0, 1))

data$education[data$AGE < 19] <- 0
data$education[data$AGE < 19 & (data$ fam_edu == 1)] <- 1

```

## Housing Quality

#### Overcrowding

Overcrowding is assigned at the household level. The individuals in the household are assigned the status of overcrowded if there are more than two people per bedroom. 

```{r overcrowding, message = FALSE}
data = data %>%
  group_by(SERIAL, YEAR) %>%
  mutate(hhsize = n())
           

data$overcrowd = if_else(data$hhsize > 2*data$BEDROOMS, 1, 0)

```


#### Housing costs

Houses are considered cost burdened at above 30% of total income and severely cost burdened at over 50% of total income. Here I follow Dhongde and Haveman 2016 in using severe cost burden as the indicator. 

```{r}
data$OWNCOST[data$OWNCOST == 99999] <- NA
data$FTOTINC[data$FTOTINC == 99999] <- NA

data$hcost <- apply(data[ ,c("OWNCOST", "RENTGRS")], 1, max, na.rm = TRUE)
data$hcost_dep <- if_else(data$hcost*12/data$FTOTINC > .50, 1, 0) # HUD uses 50% as severe cost burdened
data$hcost_dep[is.na(data$hcost_dep)] <- 0
```


## Social Connections

#### Internet/Computer Access

Internet access is assigned at the household level. A household is considered deprived if there is either no internet access, or there is none of the following computer equipment: laptop, desktop, notebook, smartphone, other computer equipment (not including GPS or household appliances). 


```{r internet}
data$computer_internet <- if_else(data$CINETHH == 3|(data$CILAPTOP == 2 & data$CIHAND == 2 & data$CIOTHCOMP == 2), 1, 0)

```


####Linguistic isolation
Linguistic isolation is calculated as not having anyone in the household over the age of 14 who speaks English. It is calculated by the census bureau. 

```{r}
data$lang_dep <- if_else(data$LINGISOL == 2, 1, 0)
```


##Subsetting the Data

Saving only the data that is used to construct the index, as well as a few groups it can be broken into (e.g. by state, gender, etc.)

```{r subset}
mpi_data = data %>%
  select(YEAR, SERIAL, PERWT, SEX, METRO, STATEFIP, income_poverty, fam_emp, health, education, overcrowd, computer_internet, lang_dep, hcost_dep, disability)
rm(data) # removing to save memory

```


##Classifying Individuals as MPI poor

Each of the five dimensions is equally weighted (.2). For the dimensions with two indicators they are equally weighted (.1) while the Education indicator has the full weight of the education dimension (.2). 

I define MPI poor as being poor in 3 or more weighted indicators, with Education being weighted at .2 because it is the only indicator in its dimension. Thus, a weighted score of over .29 is MPI poor. 

```{r mpi poor}
mpi_data$mpi_score = mpi_data$income_poverty*(1/9) + 
  mpi_data$fam_emp *(1/9) + 
  mpi_data$health*(1/6) + 
  mpi_data$education*(1/6) +
  mpi_data$overcrowd*(1/12) +
  mpi_data$computer_internet*(1/12) + 
  mpi_data$disability*(1/12) +
  mpi_data$lang_dep * (1/12) +
  mpi_data$hcost_dep * (1/9)

mpi_data$mpi_poor = if_else(mpi_data$mpi_score > .32, 1, 0)

mpi_data$mpi_score_no_inc =  
  mpi_data$fam_emp *(1/6) + 
  mpi_data$health*(1/6) + 
  mpi_data$education*(1/6) +
  mpi_data$overcrowd*(1/12) +
  mpi_data$computer_internet*(1/12) + 
  mpi_data$disability*(1/12) +
  mpi_data$lang_dep * (1/12) +
  mpi_data$hcost_dep * (1/6)

mpi_data$mpi_poor_no_inc = if_else(mpi_data$mpi_score_no_inc > .32, 1, 0)

mpi_data_15 <- mpi_data %>% filter(YEAR == 2015)
```

Writing out the data.
```{r}
write_feather(mpi_data, "C:/Users/natek/Documents/mpi_data.feather")
write_feather(mpi_data_15, "C:/Users/natek/Documents/mpi_data_15.feather")
rm(mpi_data)

```

##National Alkire-Foster Index

Here I calculate an AF index for the U.S. in 2015. 
```{r}
censored_mpi_data_15 <- mpi_data_15
censored_mpi_data_15[censored_mpi_data_15$mpi_poor == 0, 7:15] <- 0

svy_df <- svydesign(ids = ~1, weights = ~PERWT, data = mpi_data_15)
cen_svy_df <- svydesign(ids = ~1, weights = ~PERWT, data = censored_mpi_data_15)

cen_hc_tbl <- svymean(~income_poverty+fam_emp+health+disability+education+overcrowd+hcost_dep+
                           computer_internet+lang_dep+mpi_poor, design = cen_svy_df)

cen_hc_table <- ftable(cen_hc_tbl)

incidence_tbl <- svymean(~income_poverty+fam_emp+health+disability+education+overcrowd+hcost_dep+
                           computer_internet+lang_dep+mpi_poor, design = svy_df)
incidence_table <- ftable(incidence_tbl)
incidence_table <- data.frame(incidence_table[,1], incidence_table[,2], cen_hc_table[, 1], cen_hc_table[, 2])
colnames(incidence_table) <- c("Uncensored Mean", "Uncensored SE", "Censored Mean", "Censored SE")



intensity_tbl <- svyby(~mpi_score, ~I(mpi_poor == 1),design=svy_df, svymean)
intensity_table <-ftable(intensity_tbl)

headcount <- incidence_table[10,1:2]
intensity <- c(intensity_tbl$mpi_score[[2]],intensity_tbl$se[[2]])
adj_hc <- c(headcount[1]*intensity[1], "")
adjusted_hc <- adj_hc[[1]]

mpi_overview<- rbind(as.numeric(headcount), as.numeric(intensity), as.numeric(adj_hc))
row.names(mpi_overview) <- c("headcount", "intensity", "adj_hc")
colnames(mpi_overview) <- c("Value", "Standard Error")

weights <- c(.1,.1,.1,.1,.2,.1,.1,.1,.1)
per_contrib_vector <- (cen_hc_table[1:9,1]/adjusted_hc)*weights
per_contrib_df <- data.frame(per_contrib_vector)
#names(per_contrib_df) <- row.names(incidence_table)[1:9]
  
mpi_overview <- mpi_overview * 100
incidence_table <- incidence_table * 100
per_contrib_df <- per_contrib_df *100

inc_mpi_crosstab <- svytable(~income_poverty+mpi_poor, design = svy_df, Ntotal = 100)
inc_mpi_crosstab <- ftable(inc_mpi_crosstab)

no_inc_mpi_crosstab <- svytable(~income_poverty+mpi_poor_no_inc, design = svy_df, Ntotal = 100)
no_inc_mpi_crosstab <- ftable(no_inc_mpi_crosstab)

```

```{r csv}
write_csv(mpi_overview, "output_data/mpi_overview.csv")
write_csv(incidence_table, "output_data/incidence_table.csv")
write_csv(per_contrib_df, "output_data/per_contrib_df.csv")
write_csv(inc_mpi_crosstab, "output_data/inc_mpi_crosstab.csv")
write_csv(no_inc_mpi_crosstab, "output_data/no_inc_mpi_crosstab.csv")
```

```{r stargazer, eval = FALSE}
stargazer(mpi_overview, title = "MPI Overview", digits = 2, align = TRUE)
stargazer(incidence_table, summary = FALSE, title = "Incidence", digits = 2, align = TRUE)
stargazer(per_contrib_df, summary = FALSE, title = "Percent Contribution to Adjusted Headcount", 
          digits = 2, align = TRUE)
stargazer(inc_mpi_crosstab, title = "Income Poor v. MPI Poor", digits = 2, align = TRUE)
stargazer(no_inc_mpi_crosstab, title = "Income Poor v. MPI Poor (Income not included in MPI)", digits = 2, align = TRUE)

```


```{r}
library(VennDiagram)
grid.newpage()
draw.pairwise.venn(16.7, 15.5, 8.4, category = c("Income Poor", "MPI Poor"), lty = rep("blank", 
    2), fill = c("red", "blue"), alpha = rep(0.5, 2), cat.pos = c(0, 
    0), cat.dist = rep(0.025, 2))
```


finding the standard error for the MPI table
```{r}
vcov(svymean(~mpi_poor+mpi_score, cen_svy_df)) ## gives covariance of xy

vcov(svymean(~mpi_poor^2+mpi_score^2, cen_svy_df))

svymean(~mpi_poor+mpi_score, cen_svy_df)

```
```{r}
#cov of squares plus mean of x squared plus variance of x times mean of y squared plus variance of y
n <- 3147005  
expected_squares <- 2.47*10^-8 + (.164^2 + (3*10^-4)^2)/n * (.104^2 +(1*10^-4)^2)/n  
   
#mean of x times mean of y plus covariance of x and y
expected_product  <- (.164 * .104)/n + (2.47 * 10^-8)

expected_squares
expected_product
temp_ans <- sqrt(expected_squares - expected_product^2)
temp_ans
temp_ans*100
  
    
```


##Printing Results

Below I show the headcount, intensity, and adjusted headcount for the U.S. in 2015. I also show the percent of the population experiencing each indicator and the percent contribution each indicator makes to the index. 
```{r}
# headcount
# intensity
# adj_hc
# 
# col1 <- c(income_per,fam_emp_per,health_per,education_per, overcrowd_per, computer_internet_per, disability_per, 
#         lang_dep_per, hcost_dep_per)
# col1 <- round(col1, 2)
# col2 <- c(per_contrib_income, per_contrib_fam_emp, per_contrib_health, per_contrib_education, 
#           per_contrib_overcrowd, per_contrib_computer_internet, per_contrib_disability, per_contrib_lang_dep,
#           per_contrib_hcost_dep)
# col2 <- round(col2, 2)
# 
# ind_table <- cbind(col1, col2)
# rownames(ind_table) <- c("Income", "Employment", "Health", "Education", "Overcrowding", "Internet", "Disability",
#                          "Language", "Housing Cost")
# colnames(ind_table) <- c("Incidence", "Contribution")
# 
# print(ind_table)
# 
# library(stargazer)
# stargazer(ind_table[,1:2], title = "Incidence of each MPI Indicator")
# stargazer(ind_table[,c(1,3)], title = "Contribution to Adjusted Headcount of each MPI Indicator")

```

Breakdown by Metro status

```{r metro, eval = FALSE}



# metro_mpi <- mpi_data_15 %>%
#   group_by(METRO) %>%
#   summarise(headcount = sum(mpi_poor*PERWT)/sum(PERWT),
#             income_censored_hc = sum(income_poverty[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm =TRUE),
#             fam_emp_censored_hc = sum(fam_emp[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
#             health_censored_hc = sum(health[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
#             education_censored_hc = sum(education[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
#             overcrowd_censored_hc = sum(overcrowd[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
#             computer_internet_censored_hc = sum(computer_internet[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
#             disability_censored_hc = sum(disability[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
#             lang_dep_censored_hc = sum(lang_dep[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
#             hcost_dep_censored_hc = sum(hcost_dep[mpi_poor == 1]*PERWT[mpi_poor == 1], na.rm = TRUE),
# 
#             mpi_poor_hc = sum(mpi_poor*PERWT, na.rm = TRUE),
#             n = sum(PERWT),
#             income_per = sum(income_poverty * PERWT)/n,
#             fam_emp_per = sum(fam_emp * PERWT)/n,
#             health_per = sum(health * PERWT)/n,
#             education_per = sum(education * PERWT)/n,
#             overcrowd_per = sum(overcrowd * PERWT)/n,
#             computer_internet_per = sum(computer_internet * PERWT)/n,
#             disability_per = sum(disability * PERWT)/n,
#             lang_dep_per = sum(lang_dep * PERWT)/n,
#             hcost_dep_per = sum(hcost_dep * PERWT)/n)
# 
# 
# metro_mpi <- metro_mpi %>%
#   mutate(intensity = (income_censored_hc*.1 + fam_emp_censored_hc *.1 +
#            health_censored_hc * .1 + education_censored_hc * .2 +
#            overcrowd_censored_hc * .1 + computer_internet_censored_hc * .1 +
#            disability_censored_hc * .1 + lang_dep_censored_hc * .1 + hcost_dep_per * .1)/ mpi_poor_hc,
#          adj_hc = headcount*intensity,
#          per_contrib_income = ((income_censored_hc/n)/adj_hc)*.1,
#          per_contrib_fam_emp = ((fam_emp_censored_hc/n)/adj_hc)*.1,
#          per_contrib_health = ((health_censored_hc/n)/adj_hc)*.1,
#          per_contrib_education = ((education_censored_hc/n)/adj_hc)*.2,
#          per_contrib_overcrowd = ((overcrowd_censored_hc/n)/adj_hc)*.1,
#          per_contrib_computer_internet = ((computer_internet_censored_hc/n)/adj_hc)*.1,
#          per_contrib_disability = ((disability_censored_hc/n)/adj_hc)*.1,
#          per_contrib_lang_dep = ((lang_dep_censored_hc/n)/adj_hc)*.1,
#          per_contrib_hcost_dep = ((hcost_dep_censored_hc/n)/adj_hc)*.1)%>%
#   select(METRO, n, headcount, intensity, adj_hc, per_contrib_income, per_contrib_fam_emp, 
#          per_contrib_health, per_contrib_education, per_contrib_overcrowd, per_contrib_computer_internet,
#          per_contrib_disability, per_contrib_lang_dep, per_contrib_hcost_dep)
# 
# metro_mpi$METRO <- recode(metro_mpi$METRO, `0` = "Not identifiable", `1` = "Not in Metro Area", `2` = "In Metro Area, Central/Principal City", `3` = "In Metro Area,Outside Central/Principal City", `4` = "Central/Principal City status unknown")
# 



```

Breakdown by gender
```{r}
gender_tbl <- svyby(~mpi_poor+income_poverty+fam_emp+health+disability+education+overcrowd+hcost_dep+ computer_internet+lang_dep, ~I(SEX == 1),design=svy_df, svymean)

female_value_vector <- gender_tbl[1, 2:11]
male_value_vector <- gender_tbl[2, 2:11]
female_se_vector <- gender_tbl[1, 12:21]
male_se_vector <- gender_tbl[2, 12:21]
female_value_vector <- t(female_value_vector)
male_value_vector <- t(male_value_vector)
female_se_vector <- t(female_se_vector)
male_se_vector <- t(male_se_vector)


gender_df <- cbind.data.frame(female_value_vector, female_se_vector, male_value_vector, female_se_vector)
names(gender_df) <- c("Female Mean", "Female SE", "Male Mean", "Male SE")
row.names(gender_df) <- c("MPI Poor", "Income", "Employment", "Health Insurance", "Disability", "Education", "Overcrowding", "Housing Cost", "Internet", "Linguistic Isolation")

gender_tbl_cen <- svyby(~mpi_poor+income_poverty+fam_emp+health+disability+education+overcrowd+hcost_dep+ computer_internet+lang_dep, ~I(SEX == 1),design=cen_svy_df, svymean)

female_value_vector <- gender_tbl_cen[1, 2:11]
male_value_vector <- gender_tbl_cen[2, 2:11]
female_se_vector <- gender_tbl_cen[1, 12:21]
male_se_vector <- gender_tbl_cen[2, 12:21]
female_value_vector <- t(female_value_vector)
male_value_vector <- t(male_value_vector)
female_se_vector <- t(female_se_vector)
male_se_vector <- t(male_se_vector)


gender_cen_df <- cbind.data.frame(female_value_vector, female_se_vector, male_value_vector, female_se_vector)
names(gender_cen_df) <- c("Female Mean", "Female SE", "Male Mean", "Male SE")
row.names(gender_cen_df) <- c("MPI Poor", "Income", "Employment", "Health Insurance", "Disability", "Education", "Overcrowding", "Housing Cost", "Internet", "Linguistic Isolation")

per_contrib_fem <- (gender_cen_df$`Female Mean`[2:10]/adjusted_hc)*weights
per_contrib_male <- (gender_cen_df$`Male Mean`[2:10]/adjusted_hc)*weights
per_contrib_gender <- cbind.data.frame(per_contrib_fem, per_contrib_male)
names(per_contrib_gender) <- c("Female", "Male")
row.names(per_contrib_gender) <- c("Income", "Employment", "Health Insurance", "Disability", "Education", "Overcrowding", "Housing Cost", "Internet", "Linguistic Isolation")


gender_df <- gender_df*100
per_contrib_gender <- per_contrib_gender*100

```

```{r csv2}
write_csv(gender_df, "output_data/gender_df.csv")
write_csv(per_contrib_gender, "output_data/per_contrib_gender.csv")

```


```{r stargazer 2, eval = FALSE}
stargazer(gender_df, summary = FALSE, title = "Indicator Incidence by Gender", digits = 2, align = TRUE)
stargazer(per_contrib_gender, summary = FALSE, title = "Percent Contribution to Index by Gender", digits = 2, align = TRUE)
```

