---
title: "A Multidimensional Poverty Index for the U.S."
subtitle: "<br/>A Dissertation Talk"
author: "Nate Kratzer"
date: "11/30/2017"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r, include = FALSE}
library(tidyverse)
library(formattable)
library(knitr)
```

class: inverse, middle, center

# Preamble


---
# Not Quite a Job Talk

- I have job: Data Science Analyst at Brown-Forman
- But I still need faculty feedback on my academic work
- Focus is on areas of the dissertation that need the most help

---
# The Brown-Forman Job

- The second round interview was a job talk on a sample analytics project
  + I was given one hour.
  + I presented for 8 minutes. 

- Two projects so far
  + Sustainability of White Oak Trees
  + Demographic Data for Sales
  
---

# Dissertation Schedule

- I have a plan for finishing with a full-time job
  + 10 hours on weekends (5 hrs each morning)
  + 4.5 hours on evenings (3 evenings of 1.5 hrs)
- The plan includes public accountability
  + Chapters 1 (Theory) and 3 (Index Construction) revised and done by end of Dec.
  + Chapter 2 (Lit Review) by end of Jan.
  + Chapter 4 (Index Validation) by end of Feb.
  + Chapter 5 (Index Usefulness) by April 21
- Feel free to ask me if I'm on schedule and publicly shame me if I'm not. 
  

---
class: inverse, middle, center

# A Multidimensional Poverty Index for the U.S.

---

# Motivation

- The current poverty line is terrible
  + It has no defenders, the only debate is how to replace it.
- International Development research has suggested a better way 
  + A multidimensional poverty index
- The goal of this project is a little different than a typical paper
  + I am not trying to establish causal relationships
  + I am trying to measure something better to aid in future research
  + Closer to building a microscope than conducting research with a microscope
  + Goal is a clearer vision of the actual experience of poverty/deprivation
  + Final dissertation chapter gives examples of use

---

# Problems with the poverty line

- Outdated
- No geographical adjustment
- Only counts income


---

# Goals of Poverty Measurement

- Capture what we intuitively mean when we talk about individuals/families in poverty. 
  + Experiencing substantive deprivations in one or more areas of life.
- Neutral about how resources are provided (e.g. purchased, in-kind charitable, government provided, etc.)
- Measures depth of poverty
  + Welfare Reform in 1996 increased deep poverty

---

# Normative Claims in Constructing the Index

- I am claiming that each measure in index is valuable in and of itself
- Example: Education is valuable apart from its role in increasing income
  + If you don't believe this, then you should place zero weight on the education indicator
- The current normative claim embodied in the official poverty measure is that other things are only valuable if they lead to increased income
  + Consider it an index where all the weight is placed on the income category
  + Preview: I don't have a perfect weighting scheme, but can do better than placing 100% on income

---

# Normative Claims and Information

- "...the priorities that are accepted, often implicitly, in the different approaches to ethics, welfare economics, and political philosophy can be brought out and analyzed through identifying the information on which the evaluative judgments rely" (Sen 1999, p. 85)

- Liberty for libertarians
- Equality for egalitarians
- Utility for utilitarians
- Primary Goods for Rawlsians
- Capabilities for the MPI

---

# The Capabilities Approach to Measurement

- Measure _ends_ not _means_
- The goal is flourishing (Eudaimonia)
- Capabilities are a measurement of human freedom to achieve various lifestyles
  + Implies positive freedom and not solely freedom from coercion
- Human beings have widely varying tastes, but there is a minimum standard of basic capabilities
  + Ex: Food distribution and famine
  
---

# Multidimensional Poverty Indices in International Development

- Oxford Poverty and Human Development Initiative

---

# MPIs in the U.S.

---

# Structure of Proposed MPI

1. Economic (1/3)
    + Income Poverty (1/3)
    + Employment (1/3)
    + Housing Costs (1/3)
2. Educational (1/3)
    + Educational Attainment (1/2)
    + Internet Access (1/4)
    + Linguistic Isolation (1/4)
3. Health (1/3)
    + Insurance (1/2)
    + Disability (1/4)
    + Overcrowding (1/4)
    
- Division into three dimensions based on Multiple Correspondence Analysis and Interpretability
    
---

# Weights

- Multiple correspondence analysis
- PCA for dummies....the variables, not the analyst

---

# Data

- IPUMS American Community Survey Microdata, 2013-2015
  + Just presenting on 2015 today. 
- Index requires data at individual level to measure joint distribution of deprivations
  + This is the one of the reasons for an index instead of a dashboard. 

---

# Alkire-Foster Indexing (1)

- Each indicator has a cutoff value. The index takes a matrix of 0s and 1s, with 1 indicating deprivation
- Consider the first 10 people in the U.S. 2015 dataset:

```{r, echo = FALSE, warning = FALSE, message = FALSE}
sample_mpi_rows <- read_csv("output_data/few_rows_mpi.csv")
names(sample_mpi_rows) <- c("Inc", "Emp", "Hlth", "Edu", "Crowd", "Int", "Ling", "House", "Dis", "MPI_S", "MPI")
sample_mpi_rows$MPI_S <- round(sample_mpi_rows$MPI_S, 2)
mpi_formatter <- function() {
    formatter("span", 
        style = ~ style(color = ifelse(`MPI_S` > .33, "blue", "black"))
    )	 	 
} 
formattable(sample_mpi_rows, list(MPI_S = mpi_formatter(),
                                  MPI = mpi_formatter(),
                                  Inc = mpi_formatter(),
                                  Emp = mpi_formatter(),
                                  Hlth = mpi_formatter(),
                                  Edu = mpi_formatter(),
                                  Crowd = mpi_formatter(),
                                  Int = mpi_formatter(),
                                  Ling = mpi_formatter(),
                                  House = mpi_formatter(),
                                  Dis = mpi_formatter()))
```

---

# A-F Indexing (2)

- Two parts to poverty measurement
  + Identification
  + Aggregation
- A-F Aggregation
  + Headcount * Intensity
  + Insensity is the mean deprivation score among the poor
  + In previous table
    + Four of 10 were poor, so headcount = 0.4
    + Poverty scores were (.36, .36, .47, .36), intensity = `r c(.36, .36, .47, .36) %>% mean() %>% round(2)`

---

# U.S. Results

```{r, echo = FALSE, message = FALSE, warning = FALSE}
inc_mpi_crosstab <- read_csv("output_data/inc_mpi_crosstab.csv")

inc_mpi_col1 <- unlist(unname(c(inc_mpi_crosstab[4,3], inc_mpi_crosstab[2,3])))
inc_mpi_col2 <- unlist(unname(c(inc_mpi_crosstab[3,3], inc_mpi_crosstab[1,3])))
inc_mpi_summary <- data_frame(inc_mpi_col1, inc_mpi_col2)
names(inc_mpi_summary) <- c("Income Poor", "Not Income Poor")
row.names(inc_mpi_summary) <- c("MPI Poor", "Not MPI Poor")
inc_mpi_summary <- round(inc_mpi_summary, 2)
```

```{r, echo = FALSE, message = FALSE}
formattable(inc_mpi_summary)
```

---

# Incidence of Each Indicator
```{r, echo = FALSE, message = FALSE, warning = FALSE}
incidence_table <- read_csv("output_data/incidence_table.csv") 
incidence_table <- round(incidence_table, 2)
row.names(incidence_table) <- c("Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation", "MPI")
formattable(incidence_table, list(`Uncensored Mean` = color_bar('lightblue', na.rm = TRUE),
                                  `Censored Mean` = color_bar('lightblue', na.rm = TRUE)))
```

---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
per_contrib_df <- read_csv("output_data/per_contrib_df.csv") %>% round(2)
row.names(per_contrib_df) <- c("Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation")
names(per_contrib_df) <- c("Percent Contribution to Index")
formattable(per_contrib_df, list(`Percent Contribution to Index` = color_bar('lightblue')))
```

---
```{r, echo = FALSE, message = FALSE, warning = FALSE}
gender_df <- read_csv("output_data/gender_df.csv") %>% round(2)
row.names(gender_df) <- c("Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation", "MPI")
formattable(gender_df, list(`Female Mean` = color_bar('lightblue', na.rm = TRUE),
                                  `Male Mean` = color_bar('lightblue', na.rm = TRUE)))

```


---

#Percent Contribution of Each Indicator
```{r, echo = FALSE, message = FALSE, warning = FALSE}
per_contrib_gender <- read_csv("output_data/per_contrib_gender.csv") %>% round(2)
row.names(per_contrib_gender) <- c("Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation")
names(per_contrib_df) <- c("Percent Contribution to Index")
fgreater_formatter <- function() {
    formatter("span", 
        style = ~ style(color = ifelse(`Female` > `Male`, "blue", "black"))
    )	 	 
} 
mgreater_formatter <- function() {
    formatter("span", 
        style = ~ style(color = ifelse(`Male` > `Female`, "blue", "black"))
    )	 	 
} 
formattable(per_contrib_gender, list(`Male` = mgreater_formatter(),
                                 `Female` = fgreater_formatter()))
```

---

# MPI with Income indicator excluded
```{r, echo = FALSE, message = FALSE, warning = FALSE}
no_inc_mpi_crosstab <- read_csv("output_data/no_inc_mpi_crosstab.csv")
no_inc_mpi_col1 <- unlist(unname(c(no_inc_mpi_crosstab[4,3], no_inc_mpi_crosstab[2,3])))
no_inc_mpi_col2 <- unlist(unname(c(no_inc_mpi_crosstab[3,3], no_inc_mpi_crosstab[1,3])))
no_inc_mpi_summary <- data_frame(no_inc_mpi_col1, no_inc_mpi_col2)
names(no_inc_mpi_summary) <- c("Income Poor", "Not Income Poor")
row.names(no_inc_mpi_summary) <- c("MPI Poor", "Not MPI Poor")
no_inc_mpi_summary <- round(no_inc_mpi_summary, 2)
formattable(no_inc_mpi_summary)
```


---

# State Results

---

# Results by Gender

