---
title: "A Multidimensional Poverty Index for the U.S."
subtitle: "<br/>A Dissertation Talk"
author: "Nate Kratzer"
date: "11/30/2017"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r, include = FALSE}
library(tidyverse)
library(formattable)
library(knitr)
library(leaflet)
library(rgdal)
```

class: inverse, middle, center

# Preamble


---
# Not Quite a Job Talk

- I have job: Data Science Analyst at Brown-Forman
- But I still need faculty feedback on my academic work
- Focus is on areas of the dissertation that need the most help

---
# The Brown-Forman Job

- The second round interview was a job talk on a sample analytics project
  + I was given one hour.
  + I presented for 8 minutes. 

- Two projects so far
  + Sustainability of White Oak Trees
  + Demographic Data for Sales
  
---

# Dissertation Schedule

- I have a plan for finishing with a full-time job
  + 10 hours on weekends (5 hrs each morning)
  + 4.5 hours on evenings (3 evenings of 1.5 hrs)
- The plan includes public accountability
  + Chapters 1 (Theory) and 3 (Index Construction) revised and done by end of Dec.
  + Chapter 2 (Lit Review) by end of Jan.
  + Chapter 4 (Index Validation) by end of Feb.
  + Chapter 5 (Index Usefulness) by April 21
- Feel free to ask me if I'm on schedule and publicly shame me if I'm not. 
  

---
class: inverse, middle, center

# A Multidimensional Poverty Index for the U.S.

---

# Dissertation Outline

- 5 Chapters
  1. Theory
  2. Lit Review
  3. Index
  4. Validation
  5. Applications
  
I am mainly focused on 1, 3, and 5 in this presentation

---

# Motivation

- The current poverty line is terrible
  + It has no defenders, the only debate is how to replace it.
- International Development research has suggested a better way 
  + A multidimensional poverty index
- The goal of this project is a little different than a typical paper
  + I am not trying to establish causal relationships
  + I am trying to measure something better to aid in future research
  + Closer to building a microscope than conducting research with a microscope
  + Goal is a clearer vision of the actual experience of poverty/deprivation
  + Final dissertation chapter gives examples of use

---

# Problems with the poverty line

- Outdated
- No geographical adjustment
- Only counts income


---

# Goals of Poverty Measurement

- Capture what we intuitively mean when we talk about individuals/families in poverty. 
  + Experiencing substantive deprivations in one or more areas of life.
- Neutral about how resources are provided (e.g. purchased, in-kind charitable, government provided, etc.)
- Measures depth of poverty
  + Welfare Reform in 1996 increased deep poverty. 
  + Focused benefits on the working poor, who tended to be better off and closer to moving over the poverty line.
  + From 1996 - 2011 the number of households living on under $2/day/person doubled to 1.5 million, including 3 million children (Edin and Shaefer 2015).
- Compare across subgroups and over time. 

---

# Normative Claims in Constructing the Index

- I am claiming that each measure in index is valuable in and of itself
  + Means/Ends or Hypothetical/Categorical Imperatives
- Example: Education is valuable apart from its role in increasing income
  + If you don't believe this, then you should place zero weight on the education indicator
- The current normative claim embodied in the official poverty measure is that other things are only valuable if they lead to increased income
  + Consider it an index where all the weight is placed on the income category
  + Preview: I don't have a perfect weighting scheme, but I can do better than placing 100% on income

---

# Normative Claims and Information

- "...the priorities that are accepted, often implicitly, in the different approaches to ethics, welfare economics, and political philosophy can be brought out and analyzed through identifying the information on which the evaluative judgments rely" (Sen 1999, p. 85)

- Liberty for libertarians
- Equality for egalitarians
- Utility for utilitarians
- Primary Goods for Rawlsians
- Capabilities for the MPI

---

# The Capabilities Approach to Measurement

- Measure _ends_ not _means_
- The goal is flourishing (Eudaimonia)
- Capabilities are a measurement of human freedom to achieve various lifestyles
  + Implies positive freedom and not solely freedom from coercion
- Human beings have widely varying tastes, but there is a minimum standard of basic capabilities
  + Ex: Food distribution and famine
  + Instead of asking "What is the optimal state of the world?" ask, "Is state of the world A better or worse than state of the world B?"
  
---

# Multidimensional Poverty Indices in International Development

- Oxford Poverty and Human Development Initiative
  + _Multidimensional Poverty Measurement and Analysis_ (2015)
  
- MPIs and income poverty are not well matched
- Trends of non-income and income poverty are not matched
- Economic growth does not guarantee poverty reduction
- Dashboards overlook interconnections of deprivations among individuals
---

# MPIs in the U.S.

- 7 total proposed
- Alkire-Foster 2009 as a quick example from 2004 National Health Interview Survey
- Two most influential on my project
  + Dhongde and Haveman (2016)
  + Brookings (2016)
- But both have major drawbacks I correct
  + Children: Brookings excludes children and elderly. DH include, but by definition children can't be deprived in educational attainment (must be 18+ and lack H.S. degree)
  + I propose additional indicators - internet, housing costs
  + Brookings doesn't AF index, DH does but doesn't decompose index

---

# Structure of Proposed MPI

1. Economic (1/3)
    + Income Poverty (1/3)
    + Employment (1/3)
    + Housing Costs (1/3)
2. Educational (1/3)
    + Educational Attainment (1/2)
    + Internet Access (1/4)
    + Linguistic Isolation (1/4)
3. Health (1/3)
    + Insurance (1/2)
    + Disability (1/4)
    + Overcrowding (1/4)
    
- Division into three dimensions based on Multiple Correspondence Analysis and Interpretability
    
---

# Weights

- Multiple correspondence analysis used to determine dimensions, but not weights
  + If used for weights, it would overweight rarer categories
  + Substitutes a tough normative question with an easier empirical one - but they are different questions.
- Multiple Correspondence Analysis is like PCA for dummies....the variables, not the analyst

---

# Data

- IPUMS American Community Survey Microdata, 2013-2015
  + Just presenting on 2015 today. 
- Index requires data at individual level to measure joint distribution of deprivations
  + This is the one of the reasons for an index instead of a dashboard. 
  
--- 

# Defining the indicators for children

- Indicators can be assigned to individuals (health insurance), families (income), households (employment), or mixed (education is individual for adults but household for children)

- Education
  + 19 and older are poor in this dimension with no high school degree. 
  + 18 and younger are poor if they live in a household where no one has a high school degree
- Employment
  + Someone in the household is employed unless the entire household is retired or recieving substantial nonlabor income. 

---

# Alkire-Foster Indexing (1)

- Each indicator has a cutoff value. The index takes a matrix of 0s and 1s, with 1 indicating deprivation
- Consider the first 10 people in the U.S. 2015 dataset:

```{r, echo = FALSE, warning = FALSE, message = FALSE}
sample_mpi_rows <- read_csv("output_data/few_rows_mpi.csv")
names(sample_mpi_rows) <- c("Inc", "Emp", "Hlth", "Edu", "Crowd", "Int", "Ling", "House", "Dis", "MPI_S", "MPI")
sample_mpi_rows$MPI_S <- round(sample_mpi_rows$MPI_S, 2)
mpi_formatter <- function() {
    formatter("span", 
        style = ~ style(color = ifelse(`MPI_S` > .33, "blue", "black"))
    )	 	 
} 
formattable(sample_mpi_rows, list(MPI_S = mpi_formatter(),
                                  MPI = mpi_formatter(),
                                  Inc = mpi_formatter(),
                                  Emp = mpi_formatter(),
                                  Hlth = mpi_formatter(),
                                  Edu = mpi_formatter(),
                                  Crowd = mpi_formatter(),
                                  Int = mpi_formatter(),
                                  Ling = mpi_formatter(),
                                  House = mpi_formatter(),
                                  Dis = mpi_formatter()))
```

---

# A-F Indexing (2)

- Two parts to poverty measurement
  + Identification
  + Aggregation
- A-F Aggregation
  + Adjusted Headcount = Headcount * Intensity
  + Insensity is the mean deprivation score among the poor
  + In previous table
    + Four of 10 were poor, so headcount = 0.4
    + Poverty scores were (.36, .36, .47, .36), intensity = `r c(.36, .36, .47, .36) %>% mean() %>% round(2)`
    + Adjusted Headcount: `r c(.36, .36, .47, .36) %>% mean() %>% round(2) *0.4` 

---

# Incidence of Each Indicator and MPI in U.S.
```{r, echo = FALSE, message = FALSE, warning = FALSE}
incidence_table <- read_csv("output_data/incidence_table.csv") 
incidence_table <- round(incidence_table, 2)
row.names(incidence_table) <- c("Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation", "MPI")
formattable(incidence_table, list(`Uncensored Mean` = color_bar('lightblue', na.rm = TRUE),
                                  `Censored Mean` = color_bar('lightblue', na.rm = TRUE)))
```

---

# A-F Indexing (3): Percent Contribution by Indicator

$$\text{Percent Contribution} = \frac{\text{Censored Mean}}{\text{Adjusted Headcount}} \cdot \text{Weight}$$

- Censored Mean: For all nonpoor assign all indicators a value of 0 (nondeprived)

---

# Percent contribution by indicator in the U.S. 

```{r, echo = FALSE, message = FALSE, warning = FALSE}
per_contrib_df <- read_csv("output_data/per_contrib_df.csv") %>% round(2)
row.names(per_contrib_df) <- c("Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation")
names(per_contrib_df) <- c("Percent Contribution to Index")
formattable(per_contrib_df, list(`Percent Contribution to Index` = color_bar('lightblue')))
```

---
```{r, echo = FALSE, message = FALSE, warning = FALSE}
gender_df <- read_csv("output_data/gender_df.csv") %>% round(2)
row.names(gender_df) <- c("MPI", "Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation")
formattable(gender_df, list(`Female Mean` = color_bar('lightblue', na.rm = TRUE),
                                  `Male Mean` = color_bar('lightblue', na.rm = TRUE)))

```


---

#Percent Contribution of Each Indicator
```{r, echo = FALSE, message = FALSE, warning = FALSE}
per_contrib_gender <- read_csv("output_data/per_contrib_gender.csv") %>% round(2)
row.names(per_contrib_gender) <- c("Income Poverty", "Employment", "Health", "Disability", "Education", "Overcrowd", "Housing Costs", "Internet", "Linguistic Isolation")
names(per_contrib_df) <- c("Percent Contribution to Index")
fgreater_formatter <- function() {
    formatter("span", 
        style = ~ style(color = ifelse(`Female` > `Male`, "blue", "black"))
    )	 	 
} 
mgreater_formatter <- function() {
    formatter("span", 
        style = ~ style(color = ifelse(`Male` > `Female`, "blue", "black"))
    )	 	 
} 
formattable(per_contrib_gender, list(`Male` = mgreater_formatter(),
                                 `Female` = fgreater_formatter()))
```

---

# MPI compared to Income Poverty

```{r, echo = FALSE, message = FALSE, warning = FALSE}
inc_mpi_crosstab <- read_csv("output_data/inc_mpi_crosstab.csv")

inc_mpi_col1 <- unlist(unname(c(inc_mpi_crosstab[4,3], inc_mpi_crosstab[2,3])))
inc_mpi_col2 <- unlist(unname(c(inc_mpi_crosstab[3,3], inc_mpi_crosstab[1,3])))
inc_mpi_summary <- data_frame(inc_mpi_col1, inc_mpi_col2)
names(inc_mpi_summary) <- c("Income Poor", "Not Income Poor")
row.names(inc_mpi_summary) <- c("MPI Poor", "Not MPI Poor")
inc_mpi_summary <- round(inc_mpi_summary, 2)
formattable(inc_mpi_summary)
```


---

# MPI with Income indicator excluded
```{r, echo = FALSE, message = FALSE, warning = FALSE}
no_inc_mpi_crosstab <- read_csv("output_data/no_inc_mpi_crosstab.csv")
no_inc_mpi_col1 <- unlist(unname(c(no_inc_mpi_crosstab[4,3], no_inc_mpi_crosstab[2,3])))
no_inc_mpi_col2 <- unlist(unname(c(no_inc_mpi_crosstab[3,3], no_inc_mpi_crosstab[1,3])))
no_inc_mpi_summary <- data_frame(no_inc_mpi_col1, no_inc_mpi_col2)
names(no_inc_mpi_summary) <- c("Income Poor", "Not Income Poor")
row.names(no_inc_mpi_summary) <- c("MPI Poor", "Not MPI Poor")
no_inc_mpi_summary <- round(no_inc_mpi_summary, 2)
formattable(no_inc_mpi_summary)
```


---

# State Results

```{r, echo = FALSE, message = FALSE, warning = FALSE}
state_mpi_df <- read_csv("output_data/state_mpi_df.csv")

map_states <- readOGR("C:/Users/natek/Downloads/cb_2015_us_state_20m", layer = "cb_2015_us_state_20m",
                     GDAL1_integer64_policy = TRUE, verbose = FALSE)

map_states@data <- left_join(map_states@data, state_mpi_df, by = c('NAME' = 'State'))


label_hc         <- paste("Headcount:     ", round(map_states@data$headcount*100, 1))
label_int        <- paste("Intensity:        ", round(map_states@data$intensity*100, 1))
label_adj_hc     <- paste("Adj. Headcount:", round(map_states@data$adj_hc*100, 1))


labels <- sprintf(
  "<strong>%s</strong><br/>%s<br/>%s<br/>%s",
  map_states@data$NAME, label_adj_hc, label_hc, label_int
) %>% lapply(htmltools::HTML)

map_states@data$adj_hc_per <- map_states@data$adj_hc*100

pal <- colorNumeric(
  palette = "BuPu",
  domain = map_states@data$adj_hc_per
)


leaflet(map_states) %>%
  addTiles() %>%
  addPolygons(color = "#444444", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0.5,
              fillColor = ~pal(adj_hc_per),
              label = labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"))%>%
  addLegend(pal = pal, values = ~adj_hc_per, opacity = 0.7, title = "Adj. Headcount",
          position = "bottomright") %>%
  setView(lng = -95, lat = 38.2527, zoom = 3)
  
```




